---
title: 'Reproducible Research: Peer Assessment 1'
author: "Khurram Majeed"
output:
html_document:
keep_md: yes
pdf_document: default
---

### Set options and libraries
```{r}
## Always make code visible
echo = TRUE;
## numbers >= 10^5 will be denoted in scientific notation,
## and rounded to 2 digits
options(scipen = 1, digits = 2)

# Install ggplot2 package
#install.packages("ggplot2")
## Load the library to create plots
library(ggplot2)

# Install and load mice package
#install.packages("mice")
library(mice)
```

## Loading and preprocessing the data
```{r}
## Unzip the dataset archive
unzip("activity.zip")

## Read the CSV, set the class of each variable
data = read.csv("activity.csv", 
                colClasses = c("integer", "Date", "factor"))

## Add month column for later use in histogram
data$month <- as.numeric(format(data$date, "%m"))

## Remove the missing values from the dataset
actDataComplete = data[complete.cases(data),] 

## Set name of each row in the complete dataset to be the row number
rownames(actDataComplete) <- 1:nrow(actDataComplete)

## Checking the summary of new data
summary(actDataComplete)
## Checking the structure of new data
str(actDataComplete)
```

## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
```{r}
ggplot(actDataComplete, aes(date, steps)) + 
  geom_bar(stat = "identity", colour = "violetred", fill = "violetred", width = 0.7) + 
  facet_grid(. ~ month, scales = "free") + 
  labs(title = "Histogram of total number of steps per day during October & November of 2012", 
       x = "Date", y = "Total number of steps")
```

2. Calculate and report the mean and median total number of steps taken per day
```{r}
## Get the total number of steps taken each day
stepsPerDay <- aggregate(steps ~ date, data=actDataComplete, sum)$steps;
stepsPerDay

## Get the mean of total number of steps taken per day
mean(stepsPerDay)

## Get the median of total number of steps taken per day
median(stepsPerDay)
```

## What is the average daily activity pattern?
1. Create a time series plot of 5-minute interval average number of steps taken across all days  
```{r}
## Calculate average steps taken for every 5-minute interval
avgStepsAll = aggregate(actDataComplete$steps,
                        list(interval = as.numeric(as.character(actDataComplete$interval))), 
                        FUN = "mean")
## Assign name to the means column 
names(avgStepsAll)[2] = "avgSteps"

## Create a line plot for intervals and average steps 
ggplot(avgStepsAll,aes(interval, avgSteps)) + 
  geom_line(color = "violetred", size = 0.75) + 
  labs(title = "Time Series Plot of average steps taken in every 5-minute interval", 
       x = "5-minute intervals", 
       y = "Average number of steps taken")
```

2. Which 5-minute interval contains maximum number of steps on average?
```{r}
## Gets the interval and number of steps
avgStepsAll[avgStepsAll$avgSteps == max(avgStepsAll$avgSteps), ]
```

## Imputing missing values
1. Calculate the total number of observations with missing values in dataset
```{r}
sum(is.na(data))
```

2. Devise a strategy to impute missing data
```{r}
## Show summary to find out which columns have NA values
summary(data)
```
We are going to use mice packe that has a function to impute missing values in dataset

3. Replace missing values with data
```{r}
## Get a subset of dataste for imputation. 
## The "steps" has missing values, while "month" is used because our imputation method requires a matrix  
stepsSubset = data[c("steps", "month")]

## set seed to always reproduce same results
set.seed(150)

#Impute the dataset (Be patient it may take some time..)
imputedData = complete(mice(stepsSubset))

## Replace missing values with imputed data
data$steps = imputedData$steps

## Confirm the missing values have been imputed
summary(data)
```

4(a) Make a histogram of the total number of steps taken each day
```{r}
ggplot(data, aes(date, steps)) + 
  geom_bar(stat = "identity", colour = "violetred", fill = "violetred", width = 0.7) + 
  facet_grid(. ~ month, scales = "free") + 
  labs(title = "Histogram of total number of steps per day during October & November of 2012", 
       x = "Date", y = "Total number of steps")
```

4(b). Calculate and report the mean and median total number of steps taken per day
```{r}
## Get the total number of steps taken each day
stepsPerDay2 <- aggregate(steps ~ date, data=data, sum)$steps;
stepsPerDay2

## Get the mean of total number of steps taken per day
mean(stepsPerDay2)

## Get the median of total number of steps taken per day
median(stepsPerDay2)
```

The number of observations of number of steps without missing values was `r nrow(actDataComplete)`
and after imputation we have number of onservations increased to `r nrow(data)`. Because of this increase in the number of observations, we a saw a decrease in number of mean and median.

## Are there differences in activity patterns between weekdays and weekends?
1. Add a factor variable to dataset, with two levels - "weekday" and "weekend". 
```{r}
## Extract unabbreviated weekday from the date and create a new column
data$weekdays <- factor(format(data$date, "%A"))

## Define the levels of this factor variable
levels(data$weekdays) <- list(weekday = c("Monday", "Tuesday","Wednesday","Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))

## Confirm the levels of weekdays
table(data$weekdays)
```
2. Make a panel plot for 5-minutes and average number of steps across all weekdays and weekends.
```{r}
avgStepsWeekly <- aggregate(data$steps, 
                      list(interval = as.numeric(as.character(data$interval)), 
                           weekdays = data$weekdays),
                      FUN = "mean")

names(avgStepsWeekly)[3] <- "avgSteps"

library(lattice)
xyplot(avgStepsWeekly$avgSteps ~ avgStepsWeekly$interval | avgStepsWeekly$weekdays, 
       layout = c(1, 2), 
       type = "l", 
       xlab = "Interval", 
       ylab = "Number of steps")
```